Towards automated provenance collection for experimental runs of agent-based models

### 2023-05-12 paper/paper.tex
Gary changes

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{xcolor}
\usepackage{wrapfig}

@@

\begin{document}
%
\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }

%%% GP: to discuss -- which other authors should we include here?
%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
%
\authorrunning{D. Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
\url{https://www.hutton.ac.uk/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

The first of these two problems, the automation and replicability of
experiments  may be and as a matter of course and routine is already solved by
\emph{scripting}.  In the context of computing scripting is usually a program
or sequence of instructions carried out by another program rather than the
processor itself. This means scripting languages may be run by an
operating system, for instance, rather than compiled down to machine code
running directly on a processor.  Scripting languages work in one of two ways.
Either a program is is translated to  byte code, each code represents an
instruction to a \emph{virtual machine}.  This may be thought of as a software
computer running on the computer, and has the  obvious advantage that as long
as the byte code does the same thing everywhere, then the code written in this
language should be completely agnostic as to the hardware or operating system
it is running on.  Examples of this type of approach are Python, Java, Julia,
Perl and  C\#. In particular NetLogo uses the Java virtual machine. These
languages are very popular, as not only are they portable, they do not require
the significant overhead in terms of configuration that truly compiled
languages such as C or C++ require.  Most of the compilation and linking
process, being computationally intensive and historically slow  has already
taken place in the virtual machine. A modeller can write code once and it
should run anywhere capable of running that the virtual machine for which the
code was originally written. Because of this these scripting languages tend to
be hardware and operating system agnostic (meaning the code does not 'care'
where it is run) and more importantly fast to write and share.

More immediately relevant to the problem of automation and replicability, the
other form of scripting language is a program consisting of a series of
keywords which trigger sets of conveniently grouped, parameterised
instructions. A program, known as the \emph{the shell}, in Unix or Linux
terminology interprets these keywords and their parameters to the more arcane
instruction set required by the underlying processor to perform some
computation, such as invoking an executable machine code model. These shells
are pre-compiled parsers, so were historically limited in terms of their
expressibility and vocabulary size. Hence, they are generally less expressive
than the virtual machines discussed above, but are still Turing complete. This
means the coding of complex algorithms can be achieved but is awkward and slow
because the keywords are not optimised for general purpose computation, but for
carrying out operations at level of the operating system. For example, file
control, with atomicity at the granularity of a single file can be done very
efficiently using these kinds of languages. Because the latter is effectively
utilising the operating system of the computer directly, then this grouping of
languages is generally used to control an existing computer. That is, they are
used as \emph{job control languages} (JCL). Examples of such languages are
Bash, Ash, Dash, zsh, csh, Windows Power Shell, MSDOS, etc. Beyond the fact
that scripting provides automation for large experiment sets, it also allows a
certain degree of replicability, in that you to reproduce the experiment, then
all that is required is that the scripts are rerun. However this is
repeatability rather than replication. The aim is to just to rerun the
experiment, rather than reproduce a given set of results.  We believe that
anybody who does serious agent-based modelling experimentation should at the
very least be scripting most of their experiments. In our experience the
preparation and the execution of the experiment is relatively straight-forward
to script. Post processing to the final results of data, diagrams, etc., tends
to be a little neglected in the hurry to publish results.  Scripting is usually
(although not exclusively) a text based activity, which means it is amenable to
being programmatically produced.

The second problem of recording of data, we decided on the use of a relational
database. This was selected on the basis that such technologies have a proven
track record of recording and manipulating reasonably large volumes of data.
This is an established technology. It is also portable and many proven
libraries exist for the manipulation and use of such databases. And it is this
context that the discussion of scripting languages and virtual machines becomes
relevant. The approach we utilised when manipulating this database was a
scripting language that utilised a virtual machine, specifically Python. As
mentioned this gives a portability allowing us to develop on laptops and run
the resultant code in high-performance computing environments with little, or
no reconfiguration.

Thus the solution we selected was to use a proper JCL, in this case Bash to
provide the necessary automation and replication models and call the provenance
functions, written in a popular scripting language (Python) to update a
standard relational database. In this case we selected Sqlite3 for local
development and PostgreSQL for more heavy lifting in a distributed environment
such as in high-performance infrastructure. Thus we have created a provenance
tool which is based purely on scripting tools and a standard kind of database.

\begin{wrapfigure}{l}{0.5\textwidth} 
    \includegraphics[width=0.48\textwidth]{img/schema.jpeg}
\caption{SSREPI Schema} \label{fig:schema} \end{wrapfigure}

Since the inception of this project, we have subsequently uncovered a similar
project which specifically concentrates on provenance data. This is the FAIR
data pipeline \cite{mitchell2022fair} in which the provenance is automatically
recorded by embedded code in R, C++, Java, Julia and Python.  One of the aims
was not to modify the original code, i.e. the original experiment as unchanged
as possible. That is we wanted to change the framework code rather than the
core code. The core code in the example above being the executable, pre-compiled
in Objective-C \cite{objectivec}. Additionally the support code, i.e. the code
that prepared the data, a Perl \cite{perl563} script and the code that did the
post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed
the database access in Python \cite{python368}, wrote wrapper scripts in Bash
\cite{bash4420}. This differs in approach to [ibid] which is embedded in the
code that constitutes the body of the experiment. Moreover our framework
generalises to more more source metadata (paper, data, other experiments), and
sis written with the requirements of agent-based modelling first and foremost.
That is not to say that we do not have similar ambitions to automatically place
provenance tracing code automatically embedded in experimental model code.

The rest of this paper, we describe the scripting tool we have developed  for
automatically recording metadata, which can be incorporated into the analysis
replication process, and how this  was used to regenerate some of the figures
in \cite{polhill_nonlinearities_2013}. In addition we show a simple tool we have already developed to trace data through the provenance graph, and as usual suggest additional work we would like to pursue and other ideas.

\section{Method}

One of the artefacts of the MIRACLE project \cite{parker2019final} was the
Social Simulation REplication Interface or SSREPI. This is the schema shown in
figure \ref{fig:schema}.  This schema has evolved since it original conception.
The newest version of this document may be found here
\cite{polhill2022miracle}. This is a database schema derived from the Dublin
Core \cite{weibel2000dublin}, the standard XML datatypes \cite{biron2004xml}
and the PROV-O ontologies \cite{missier2013w3c}.  The schema is designed with
agnosticisim towards the underlying database technology and as such has been
implemented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3
\cite{sqliteorg2023syntax}.

There are two important dimensions of distinction to this schema. First is
fine- versus coarse- grained metadata. Second is provenance versus workflow.
Coarse-grained metadata describes how particular files come (or came) into
being, or were (or could be) used to bring other files into being. Fine-grained
metadata describes specific values recorded in social simulation outputs. To
make the distinction concrete, suppose a simulation produces a CSV file. The
data within the CSV file are covered by fine-grained metadata, whilst the fact
that the simulation produces the CSV file is coarse-grained. Turning to the
other dimension, provenance metadata describes what actually happens (run W of
simulation X produced output file Y), whilst workflow metadata describes what
could happen (simulation X produces an output file of type Z). The distinctions
are summarised in \ref{fig:finegrain} .

\begin{wrapfigure}{r}{0.5\textwidth}
\includegraphics[width=0.5\textwidth]{img/fine-grain-vs-coarse-grain.jpeg}
\caption{Fine grain vs coarse grain provenance} \label{fig:finegrain}
\end{wrapfigure}

Each table in schema shown in fig. \ref{fig:schema} was coded as an object type
in Python. Each table row was represented as an instance of such an object.
This design approach was adopted to enforce a consistent coding methodology
across all tables.

This then just left the interface between the two scripting languages. The
design decision was taken to do domain specific processing in the Bash
scripting language. This was done for due to the repetitive and time consuming
nature of system testing: amendments in the job control language being easier
to do quickly. This is a decision that may need revisiting. The problem being
that Bash invocations are computationally and time expensive. Complex
processing should be abstracted away from the slower language, in this case job
control language to improve speed and remove what might be regarded as
inappropriate computation for such an environment. Consequently as the system
stands the commands that coordinated between Bash and Python are few and
simple, the harder processing having been pushed down into the job control
level, i.e. in the Bash scripts.

These coordinating commands are:

\begin{itemize} \item {\color{green} \ttvar{create_database.py}} - creates a
        database idempotently.  \item {\color{green} \ttvar{exists.py}} - checks
        if a particular row in a table exists.  \item {\color{green}
            \ttvar{get_value.py}} - gets any specified single value from a
            table given the primary key.  \item {\color{green}
\ttvar{get_values.py}}  - gets one or more rows given the search values. \item {\color{green}\ttvar{next_study.py}} - gets the next available and unique study number. \item
{\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
table.  \end{itemize}

Note the use of idempotent, indicating that multiple operations on
the same entity will leave it unchanged after the initial operation.  For
example multiplying something by 1 is idempotent. This allows for a less
rigorous exception criteria, for instance if a row or entity already exists.
This implies that initialisation must be performed with care, as older data
will not necessarily be destroyed but overwritten with newer values, but
also retaining any values that already exist. Again this design decision was
expedient on rapid development and may need revisiting. However the schema is
and its interaction is quite complex and this would involve critical path
analysis

So each of the Bash commands listed in table \ref{tab:commands} composes over
the above python commands to populate the SSREPI database in a consistent and
logical manner.

For the purposes of this short paper we are concentrating on a demonstration of
recording provenance. Indeed we use the example, mentioned in the introduction
of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420}
scripts to include the what we denote as \textit{primitives}.  In order to
record provenance we coded the following primitives from the SSREPI interface
definition \cite{polhill2022miracle}. As noted above these are implemented in a
mixture of Python and Bash.

\tiny \begin{longtable}{|l|p{2cm}|p{7cm}|} \hline Primitive & Type & Purpose \\
    \hline {\color{blue} \ttvar{SSREPI_require\_minimum}} & Metadata & Lower
    bound on software hardware required \\ {\color{blue}
    \ttvar{SSREPI_require\_exact}} & Metadata & Exact bound on software
    hardware required \\ {\color{blue} \ttvar{SSREPI_application}} & Provenance
    \& Metadata & specifies some executable \\ {\color{blue} \ttvar{SSREPI_me}}
    & Provenance \& Metadata & Determines executable being run or returns a
    proper reference to the executable being run. \\ {\color{blue}
    \ttvar{SSREPI_run}} & Provenance \& metadata & Blocking invocation of an
    executable which will allow the specification of all inputs, outputs and
    arguments. Creates run-time provenance information as well \\ {\color{blue}
    \ttvar{SSREPI_batch}} & Provenance \& metadata & Non blocking invocation of
    an executable which will allow the specification of all inputs, outputs and
    arguments. Creates run-time provenance information as well \\ {\color{blue}
    \ttvar{SSREPI_argument}} & Provenance & An argument type to an exectuable
    \\ {\color{blue} \ttvar{SSREPI_output}} & Provenance & An output type from
    an executable \\ {\color{blue} \ttvar{SSREPI_input}} & Provenance & An
    input type for an exectuable \\ {\color{blue}
    \ttvar{SSREPI_hutton\_person}} & Metadata & Uses our institutions databases
    to populate metadata for a given individual \\ {\color{blue}
    \ttvar{SSREPI_person}} & Metdata & Provide metadata for a particular actor
    within this system\\ {\color{blue} \ttvar{SSREPI_project}} & Metadata &
    Specifies a project which contains all studies \\ {\color{blue}
    \ttvar{SSREPI_study}} & Metadata & A set of experiments makes up a single
    study \\ {\color{blue} \ttvar{SSREPI_set}} & Metdata & Sets the default
    licence and other metadata \\ {\color{blue} \ttvar{SSREPI_involvement}} &
    Metadata & Links personnel to a study \\ {\color{blue}
    \ttvar{SSREPI_paper}} & Metadata & A paper associated with this study \\
    {\color{blue} \ttvar{SSREPI_make_tag}} & Metadata & Used for building a
    folksonomy \\ {\color{blue} \ttvar{SSREPI_tag}} & Metadata & Used to tag
    any entity with a folksonomy tag \\ {\color{blue}
    \ttvar{SSREPI_contributor}} & Metadata & A  person with some kind of
    relation to an executable or script. \\

    {\color{blue} \ttvar{SSREPI_statistical_method}} & Metadata & A statistical
    method is an approach to computing some statistics. It may be implemented
    in or as part of an application. A statistical method generates one or more
    statistical variables as its results, and may use the results of another
    statistical method in its computation. For example, computing the standard
    deviation of some data uses the mean of those data. \\ {\color{blue}
    \ttvar{SSREPI_visualisation}} & Metadata & A visualisation is the process
    of creating an image to depict one or more (typically more than one)
    values. The results of a visualisation appear in a container.\\
    {\color{blue} \ttvar{SSREPI_statistics}} & Metadata & Statistics are
    activities that compute  and populate the values of statistical rvariables.
    They operate on raw data that are retrieved from the values using a query.
    To replicate a set of statistics, the query can be rerun, selecting values
    that are pointed to by containers entries \\ {\color{blue}
    \ttvar{SSREPI_visualisation_method}} & Metadata & This describes methods
    for generating visualisations, which then may appear in the content of a
    container produced by a process running an application that implements it.
    \\

    {\color{blue} \ttvar{SSREPI_implements}} & Metadata & Links a statistical
    or visualisation method to an application \\ {\color{blue}
    \ttvar{SSREPI_parameter}} & Metadata & A Parameter is the name of a
    parameter taken by a statistical or visualisation method, used to configure
    the way it behaves. \\ {\color{blue} \ttvar{SSREPI_statistical_variable}} &
    Metadata &  A name for (one of) the result(s) of a statistical method. \\
    {\color{blue} \ttvar{SSREPI_visualisation_variable}} & Provenance \&
    Metadata & Declares a named variable of interest \\ {\color{blue}
    \ttvar{SSREPI_variable}} & Metadata & Names a variable of interest \\
    {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & Provenance \&
    Metdata & Sets an actual value for a named statistical variable \\
    {\color{blue} \ttvar{SSREPI_value}} & Provenance & Sets an actual value.
    This can be connected to any metadata value such         \\ {\color{blue}
    \ttvar{SSREPI_content}} & Metadata & Links a kind of output/input/argument
    to a variable  \\ {\color{blue} \ttvar{SSREPI_person_makes_assumption}} &
Metadata & links a person to an assumption \\ \hline \caption{Available Bash
commands for provenance and metadata gathering} \label{tab:commands}
\end{longtable}

@@

Broadly speaking {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
\ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
\ttvar{SSREPI_input}}, {\color{blue} \ttvar{SSREPI_output}} and {\color{blue}
\ttvar{SSREPI_argument}} are responsible for recording coarse grain provenance.
{\color{blue} \ttvar{SSREPI_value}}, {\color{blue}
\ttvar{SSREPI_visualisation_variable_value}}, {\color{blue}
\ttvar{SSREPI_statistical_variable_value}}, {\color{blue} \ttvar{SSREPI_run}}
and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
remaining primitives are largely about recording metadata.

With the tool in the current state this involved laboriously going through the
code and inserting the some of the Bash-based commands, shown in
table \ref{tab:commands}. These commands have a fairly complex parameter set as
it stands, and no way of checking whether the code worked other than running
it, initially on a small subset of the experiment and then scaling up. This was
extremely time-consuming.

Elaboration of all this may be found in the man pages and design documents in
the public repository at \url{https://github.com:/DougSalt/ABM-metadata}.

\begin{figure} \includegraphics[width=\textwidth]{img/workflow.pdf}
\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}

We developed a very basic scheduler for use in an high performance computing
environment, where there are many resources available both on the local
machine. In particular for the situation where they is a single machine that
has a large memory and many cores, the non-blocking version of the  two
execution primitives, specifically {\color{blue} \ttvar{SSREPI_batch}}
implements a kind of a simple scheduler locally using the sub-processing
commands available to Unix-like systems. This was an historical solution to
schedule in an high performance computing environment that existed prior to
clustered computing situations. Clustering is the utilisation of one or more
instances of such powerful computing machinery. Generally schedulers have been
developed for such environments, as such SGE \cite{gentzsch2001sun} and the one
we use as default, Slurm \cite{yoo2003slurm}. Scheduling is now largely handled
outwith the framework, which makes the code less complex.

{\color{blue} \ttvar{SSREPI_run}}, the other executing primitive will block
until the command, code or model is has invoked has completed. This is
necessarily used in single threaded coordination context, such as gathering all
outputs necessary before proceeding to the next stage of processing. For
instance, initial post processing of an all experiments would represent such a
single threaded context.

\section{Results}

The primary result is that the target results and diagrams from the original
paper \cite{polhill2017lessons} were able to be recreated.  As mentioned the
method undertaken in this paper involved inserting the correct Bash-based
commands, listed in table \ref{tab:commands}. These commands have a fairly
complex parameter set as it stands, and no way of checking whether the code
worked other than running it. This led to one of the immediate problems with
this method. The only way to see if the code was working was to run it, a suck
it and see methodology. This was extremely time consuming. We took the obvious
approach and ran small subsets of the ~20,000 runs or so, to see if we could
get the framework to work and then scaled it up to the full run, but invariably
we ran into new problems the more we increased the scale. It therefore took
substantial time to even get this first replication to run. It took several
months of coding effort to get this correct. Generally it would alternate
between parameter errors for the calls in table \ref{tab:commands}, and errors in the scripts making calls to record the provenance data to the database. Each iteration would take weeks to run, before failure as we increased the scale.

After a run had completed successfully the database was run through several
programs to produce the sub-graphs detailed in fig. \ref{fig:schema}. These
programs are capable of producing the following subgraphs:

@@

We developed programs that took the relational data in order to produce "dot"
files. Such files that can be processed by Graphviz
\cite{ellson2002graphviz}, and were used to produce the diagrams in figures
\ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
\ref{fig:high-lit-provenance-trace} .

To give an idea of the visualisation available we show the workflow graph in
fig. \ref{fig:workflow}. This is a section of the workflow and is relative easy
to follow. The resultant provenance graph is both massive and massively
complex, given there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
\ref{fig:sub-provenance}.

\begin{figure}
\includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
\caption{A small sub-section of the proevenance graph} \label{fig:sub-provenance}
\end{figure}

So what use is this provenance metadata? Beyond replicability? Say for instance
we have a bad dataset. The challenge is not only how to visualise the
provenance, but to use this data in a useful manner. We propose a simple use
here.  The "dot" language used by Graphviz constitutes a primitive graph
database. Indeed there are programs that can transform "dot" files into
TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. Rather than
using the relational database data we used the resultant dot files listed above
to trace bad entities. \footnote{These are directed graphs (unproven as to
whether they are acyclic - they shouldn't be they are the result of a
relational database which was designed to be and \textit{should} be acyclic).}

So for an easy to see example we pick the entity {\color{red}\ttvar{Applications.application_3831436655}}, and see how this affects other entities in the graph. This is \ttvar{SSS-StopC2-Cluster-create.pl} and we are using the workflow graph to see what might be affected. And this is shown in \ref{fig:broken-application}. It might be for instance that there is a bug in this program, so we should like to see how the errors might propagate out through the workflow.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{longtable}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{wrapfig}

@@

\begin{document}
%
\title{Towards automated provenance collection for experimental runs of agent-based models}

%\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
% \author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
% Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }

%%% GP: to discuss -- which other authors should we include here?
%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
%
%\authorrunning{D. Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
% \institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
% \email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
%\url{https://www.hutton.ac.uk/}
%}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

%%% GP: Restarting 2023-05-11T15:36 -- these details on scripts can go and I am about to delete them

The automation of experiments is already solved by
\textit{scripting}. A script -- in a scripting language such as Python, Julia, or
Perl, but more typically an operating system `shell' language such as Bash --
records the sequence of programs that need to be executed to reproduce the experiment.
This at least repeats the
experiment, rather than \textit{precisely} replicating a specific set of results.
We believe that
anybody who does experiments with their agent-based models should at the
very least be scripting most of their experiments. In our experience the
preparation and the execution of the experiment is relatively straightforward
to script. Post-processing to the final data tables, diagrams, etc., that appear in
manuscripts and other documents should be included in these scripts. These scripts
can be conceived as \textit{workflow} metadata -- the means by which a \textit{class}
of outputs is achieved, rather than a specific instance.\footnote{In the case that all of the programs executed are deterministic, the class is a singleton.}

\textit{Provenance} metadata is used to record the means by which a specific output is achieved, which is more suitably stored in a 
database. That database is updated using a series of scripts written in Python, allowing us to develop on laptops and run
the resultant code in high-performance computing environments with little or
no reconfiguration. Bash scripts automating the experiment workflow are then modified to
call the Python scripts to update the provenance database. We implemented the database in Sqlite3 for local
development and PostgreSQL in high-performance infrastructure. In this way, we have created a provenance
tool that can record provenance during workflow execution.

\begin{figure} % wrapfigure is too small to see anything useful
    \includegraphics[width=\textwidth]{img/schema.jpeg} % You should use a PNG at >= 300 dpi; not a lossy format like JPEG. You can export PNGs from PDFs using Preview
\caption{SSREPI Schema} \label{fig:schema} \end{figure}

Another project adopting a similar approach is the \texttt{fair} data pipeline \cite{mitchell2022fair}, in which the provenance is automatically
recorded by embedded model code in R, C++, Java, Julia and Python. However, one of our requirements
was not to modify the model code if this could be avoided.\footnote{If the model code did not output the seed it used, then it would need to be modified.} In the example above, the FEARLUS-SPOMM model is written
in Objective-C \cite{objectivec}; with additional supporting Perl and R scripts to prepare the data and conduct the
post analysis. We would also prefer not to modify these scripts. Instead, we wrote wrapper scripts in Bash
\cite{bash4420}. A further observation is that our framework
generalises to more source metadata (papers, data, other experiments), and
is written with the requirements of agent-based modelling in mind first and foremost.

In the rest of this paper, we describe the scripting tool we have developed  for
automatically recording metadata, which can be incorporated into the analysis
replication process, and how this  was used to regenerate some of the figures
in \cite{polhill_nonlinearities_2013}. In addition we show a simple tool we have already developed to trace data through the provenance graph, and suggest additional work we would like to pursue.

\section{The \texttt{SSREPI} provenance and workflow tool}

One of the artefacts of the MIRACLE project \cite{parker2019final} was the
Social Simulation REplication Interface or \texttt{SSREPI}. This is the schema shown in
figure \ref{fig:schema}.  This schema has evolved since it original conception; 
\cite{polhill2022miracle} is the latest version. The schema is derived from the Dublin
Core \cite{weibel2000dublin} and PROV-O ontologies \cite{missier2013w3c} among others, and is designed with
agnosticism towards the underlying database technology, having been
implemented in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3
\cite{sqliteorg2023syntax}.

Besides the distinction between workflow and provenance described above, we can discriminate
fine- versus coarse- grained metadata. Coarse-grained metadata describes how particular files come (or came) into
being, or were (or could be) used to bring other files into being. Fine-grained
metadata describes specific values recorded in social simulation outputs. To
make the distinction concrete, suppose a simulation produces a CSV file. The
data within the CSV file are covered by fine-grained metadata, whilst the fact
that the simulation produces the CSV file is coarse-grained.

To build \texttt{SSREPI}, each table in schema shown in fig. \ref{fig:schema} was coded as an object type
in Python. Each table row was represented as an instance of such an object.
This design approach was adopted to enforce a consistent coding methodology
across all tables. To these were added a few simple coordinating commands that could be called from Bash:

\begin{itemize}
    \item {\color{OliveGreen} \ttvar{create_database.py}} - creates a
        database idempotently.
    \item {\color{OliveGreen} \ttvar{exists.py}} - checks
        if a particular row in a table exists.
    \item {\color{OliveGreen}
            \ttvar{get_value.py}} - gets any specified single value from a
            table given the primary key.
    \item {\color{OliveGreen} % 'green' hurt my eyes!
\ttvar{get_values.py}}  - gets one or more rows given the search values.
    \item {\color{OliveGreen}\ttvar{next_study.py}} - gets the next available and unique study number. \item
{\color{OliveGreen} \ttvar{update.py}} - idempotently updates a particular row in a
table.  \end{itemize}

`Idempotent' indicates that multiple operations on
the same entity will leave it unchanged after the initial operation. For
example multiplying something by 1 is idempotent. This allows for less
rigorous exception criteria, but implies that initialisation must be performed with care, as older data
will not necessarily be destroyed when overwritten with newer values, but
retain any values that already exist.\footnote{This is a design decision that may need revisiting.}

\begin{table}\begin{tabular}{|l|p{1cm}|p{5.5cm}|} \hline Primitive & Type & Purpose \\
    \hline {\color{blue} \ttvar{SSREPI_require_minimum}} & M & Lower
    bound on software hardware required \\ {\color{blue}
    \ttvar{SSREPI_require_exact}} & M & Exact bound on software
    hardware required \\ {\color{blue} \ttvar{SSREPI_application}} & P
    \& M & specifies some executable \\ {\color{blue} \ttvar{SSREPI_me}}
    & P \& M & Determines executable being run or returns a
    proper reference to the executable being run. \\ {\color{blue}
    \ttvar{SSREPI_argument}} & P & An argument type to an executable
    \\ {\color{blue} \ttvar{SSREPI_output}} & P & An output type from
    an executable \\ {\color{blue} \ttvar{SSREPI_input}} & P & An
    input type for an executable \\ {\color{blue}
    \ttvar{SSREPI_person}} & M & Provide metadata for a particular actor
    within this system\\ {\color{blue} \ttvar{SSREPI_project}} & M &
    Specifies a project which contains all studies \\ {\color{blue}
    \ttvar{SSREPI_study}} & M & A set of experiments makes up a single
    study \\ {\color{blue} \ttvar{SSREPI_set}} & M & Sets the default
    licence and other metadata \\ {\color{blue} \ttvar{SSREPI_involvement}} &
    M & Links personnel to a study \\ {\color{blue}
    \ttvar{SSREPI_paper}} & M & A paper associated with this study \\
    {\color{blue} \ttvar{SSREPI_make_tag}} & M & Used for building a
    folksonomy \\ {\color{blue} \ttvar{SSREPI_tag}} & M & Used to tag
    any entity with a folksonomy tag \\ {\color{blue}
    \ttvar{SSREPI_contributor}} & M & A  person with some kind of
    relation to an executable or script. \\

    {\color{blue} \ttvar{SSREPI_statistical_method}} & M & Record a statistical method \\ {\color{blue}
    \ttvar{SSREPI_visualisation}} & M & Record a method to create an image to depict one or more 
    values. \\
    {\color{blue} \ttvar{SSREPI_statistics}} & M & Record
    activities that compute and populate the values of statistical variables. \\ {\color{blue}
    \ttvar{SSREPI_visualisation_method}} & M & Methods
    for generating visualisations.
    \\

    {\color{blue} \ttvar{SSREPI_implements}} & M & Links a statistical
    or visualisation method to an application \\ {\color{blue}
    \ttvar{SSREPI_parameter}} & M & Record the name of a
    parameter taken by a statistical or visualisation method. \\ {\color{blue} \ttvar{SSREPI_statistical_variable}} &
    M &  A name for (one of) the result(s) of a statistical method. \\
    {\color{blue} \ttvar{SSREPI_visualisation_variable}} & P \&
    M & Declares a named variable of interest \\ {\color{blue}
    \ttvar{SSREPI_variable}} & M & Names a variable of interest \\
    {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & P \&
    M & Sets an actual value for a named statistical variable \\
    {\color{blue} \ttvar{SSREPI_value}} & P & Sets a value.         \\ {\color{blue}
    \ttvar{SSREPI_content}} & M & Links a kind of output/input/argument
    to a variable  \\ {\color{blue} \ttvar{SSREPI_person_makes_assumption}} &
M & Links a person to an assumption \\ \hline \end{tabular} \caption{\texttt{SSREPI}'s Bash
commands for provenance (Type column entry `P') and other metadata (Type `M') recording} \label{tab:commands}
\end{table}

@@

Each of the Bash commands listed in table \ref{tab:commands} composes over % What does 'composes over' mean?
the above Python commands to populate the \texttt{SSREPI} database in a consistent and
logical manner. Broadly speaking, {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
\ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
\ttvar{SSREPI_input}}, {\color{blue} \ttvar{SSREPI_output}} and {\color{blue}
\ttvar{SSREPI_argument}} are responsible for recording coarse grain provenance.
{\color{blue} \ttvar{SSREPI_value}}, {\color{blue}
\ttvar{SSREPI_visualisation_variable_value}}, {\color{blue}
\ttvar{SSREPI_statistical_variable_value}}, {\color{blue} \ttvar{SSREPI_run}}
and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
remaining primitives are largely about recording other metadata. Further information may be found
in the tool's public repository \footnote{\url{https://github.com/DougSalt/ABM-metadata}}.

Using \texttt{SSREPI} for existing workflow scripts is a somewhat laborious task entailing adding calls to the provenance metadata scripts in Table \ref{tab:commands}. Improvements to the `interface' is the subject of future work.

\section{Demonstration}

Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in the
paper \cite{polhill2017lessons}, showing some of the workflow and provenance graphs, and then using fictitious examples to apply the workflow and provenance metadata to addressing issues in large-scale experiments. From a workflow perspective, the main result is that the
principal diagrams in that paper were successfully recreated, with the same, albeit
not identical, results. %right? % Plus, I've deleted all the whining :-)

After a run had completed successfully the database was run through several
supplementary programs to produce the following visualisations of the provenance metadata recorded:

@@

The supplementary programs use the Dot language for input to Graphviz\cite{ellson2002graphviz}, producing the diagrams such as those in figures
\ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
\ref{fig:high-lit-provenance-trace}. The workflow graph is in
fig. \ref{fig:workflow}. The resultant provenance graph is massive, since there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
\ref{fig:sub-provenance}.

\begin{figure} % I suggest you put 'dir = LR' in this DOT file -- we can't really read the workflow
\includegraphics[width=\textwidth]{img/workflow.pdf}
\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
\caption{A small sub-section of the provenance graph} \label{fig:sub-provenance}
\end{figure}

So what use is this provenance metadata? Imagine for the purposes of illustration that
we have found a bug in a script. The entity {\color{red}\ttvar{Applications.application_3831436655}} is an experiment setup script \ttvar{SSS-StopC2-Cluster-create.pl}. We can use the workflow graph to see what else in the workflow might be affected. As might be expected for an experiment setup script, there are serious cascading consequences, which we can visualise in Figure \ref{fig:broken-application}. The Dot language used by Graphviz constitutes a primitive graph
database. Indeed there are programs that can transform Dot files into
TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. To generate Figure \ref{fig:broken-application}, we used the workflow visualisation Dot file rather than the relational database.

\begin{figure} \includegraphics[width=\textwidth]{img/broken-application.pdf}
\caption{The workflow sub-graph show the propagation of a broken application (in red) to anything transitively making use of its outputs} \label{fig:broken-application} \end{figure}
